The promise of community-driven preprints in ecology and evolution

### 2024-05-31 ms/ms.qmd
Intro

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While the adoption of preprints in biology have been slower relative to other disciplines [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed and visibility of research findings which can lead to more citations [e.g., @colavizza2024analysis]; 2) helping establish precedence of research findings; 3) removing financial barriers to open access publication; and 4) enabling feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating]. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally [@ni2024preprint].

Preprint servers can empower researchers to make their research findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally, from a lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution. Such an understanding can help inform future initiatives and allow us to work harder in further breaking down barriers to scientific publishing.

*EcoEvoRxiv* is one of the few community-driven preprint servers that has paved the way for new initiatives, by accepting multilingual preprints, recognising registered reports and non-traditional research reports, and promoting peer review and community discussion which we hope will improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI [@guillemaud2019peer]). Peer Community In allows for fast constructive peer review around a preprint. Peer reviews are transparent and published online and the community makes a recommendation on where the reasearch findings should be published [@guillemaud2019peer].

Here, we explore the first 1000+ preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We aim to better understand: 1) in what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender are associated with preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.

>>>>

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. Preprints in biology have had a slower uptake relative to other disciplines [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the visibility of research and the speed at which research findings become available which can lead to more citations [e.g., @colavizza2024analysis]; 2) helping establish precedence of research findings; 3) removing financial barriers to open access publication; and 4) enabling feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating]. Preprints increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of scientific knowledge that can have significant impacts on fundamental and applied knowledge globally [@ni2024preprint].

Preprint servers can empower researchers to make their research findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. However, significant barriers to the widespread adoption of preprints remain, ranging from a lack of clarity around preprint policies in journals [@Klebel2020] to a stigma within the research community that preprints are of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack an understanding of the factors that influence preprint use in ecology and evolution. Such an understanding may help improve current initiatives (see below), inform future ones and allow us to work harder in further breaking down barriers to scientific publishing.

*EcoEvoRxiv* is one of the few community-driven preprint servers that has paved the way for new initiatives, by accepting multilingual preprints, registered reports, and non-traditional research reports. *EcoEvoRxiv* also promotes peer review and community discussion in the hopes of improving the quality of preprints and speeding up their peer-reviewed publication (e.g., Peer Community In -- PCI [@guillemaud2019peer]). Peer Community In allows for fast constructive peer review around a preprint. Peer reviews are transparent and published online and the community makes a recommendation on where the reasearch findings should be published [@guillemaud2019peer].

Here, we explore the first 1000+ preprints uploaded to *EcoEvoRxiv* to understand preprint practices and use in ecology and evolution. We aim to understand: 1) in what countries authors who use *EcoEvoRxiv* are located; 2) the taxonomic diversity across preprints; 3) whether preprint server use depends on career stage and gender; 4) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 5) the extent to which data and code are shared in preprints; and 6) how many preprints remain unpublished, and for those that are published, how long it took for them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further clarify the benefits of using community-driven preprint servers to disseminate their research findings.
